---
title: "Cyclistic Case Study"
author: "Chad Sisk"
date: '2022-06-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/chadsisk/Education/Coursera/Google/DataAnalytics/Case Studies/Cyclistic/CSV Files/')
```

# Divvy_Exercise_Full_Year_Analysis


-- Chad Sisk  
-- Google Analytics Professional Certificate  
-- Case Study 1: Cyclistic Bikeshare Program  


This analysis is based on the Divvy case study *Sophisticated, Clear, and Polished’: Divvy and Data Visualization* written by Kevin Hartman ([found here]( https://artscience.blog/home/divvy-dataviz-case-study). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: *In what ways do members and casual riders use Divvy bikes differently?*  


**Note:** For my own case study, I used monthly data from June 2021 through May 2022. The instructions for the script given below were originally for the quarterly data. I have made modification to the script to fit the data I have chosen for my project. - Chad  


* Install required packages  
* tidyverse for data import and wrangling  
* lubridate for date functions  
* ggplot for visualization  

```{r install packages}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(rmarkdown)
getwd() #displays your working directory
setwd("/Users/chadsisk/Education/Coursera/Google/DataAnalytics/Case Studies/Cyclistic/CSV Files/") #sets your working directory to simplify calls to data
```


## STEP 1: COLLECT DATA

* Upload Divvy datasets (csv files) here  
*I uploaded monthly data instead - Chad*  

```{r collect data}
june_2021 <- read_csv("202106-divvy-tripdata.csv")
july_2021 <- read_csv("202107-divvy-tripdata.csv")
aug_2021 <- read_csv("202108-divvy-tripdata.csv")
sep_2021 <- read_csv("202109-divvy-tripdata.csv")
oct_2021 <- read_csv("202110-divvy-tripdata.csv")
nov_2021 <- read_csv("202111-divvy-tripdata.csv")
dec_2021 <- read_csv("202112-divvy-tripdata.csv")
jan_2022 <- read_csv("202201-divvy-tripdata.csv")
feb_2022 <- read_csv("202202-divvy-tripdata.csv")
mar_2022 <- read_csv("202203-divvy-tripdata.csv")
apr_2022 <- read_csv("202204-divvy-tripdata.csv")
may_2022 <- read_csv("202205-divvy-tripdata.csv")
```


## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE  

* Compare column names each of the files  
* While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file  

```{r wrangle data}
colnames(june_2021)
colnames(july_2021)
colnames(aug_2021)
colnames(sep_2021)
colnames(oct_2021)
colnames(nov_2021)
colnames(dec_2021)
colnames(jan_2022)
colnames(feb_2022)
colnames(mar_2022)
colnames(apr_2022)
colnames(may_2022)
```

-- Inspect the dataframes and look for incongruencies  

```{r}
str(june_2021)
str(july_2021)
str(aug_2021)
str(sep_2021)
str(oct_2021)
str(nov_2021)
str(dec_2021)
str(jan_2022)
str(feb_2022)
str(mar_2022)
str(apr_2022)
str(may_2022)
```

-- Convert ride_id and rideable_type to character so that they can stack correctly  
*Again using monthly data from June 2021 through May 2022, not the quarerly data from 2019 as used in the provided script -Chad*  

```{r convert data}
june_2021 <-  mutate(june_2021, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
july_2021 <-  mutate(july_2021, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
aug_2021 <-  mutate(aug_2021, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
sep_2021 <-  mutate(sep_2021, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
oct_2021 <-  mutate(oct_2021, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
nov_2021 <-  mutate(nov_2021, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
dec_2021 <-  mutate(dec_2021, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
jan_2022 <-  mutate(jan_2022, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
feb_2022 <-  mutate(feb_2022, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
mar_2022 <-  mutate(mar_2022, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
apr_2022 <-  mutate(apr_2022, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
may_2022 <-  mutate(may_2022, ride_id = as.character(ride_id)
                     ,rideable_type = as.character(rideable_type))
```

-- Stack individual quarter's data frames into one big data frame  
```{r stacking frames}
all_trips <- bind_rows(june_2021, july_2021, aug_2021, sep_2021, oct_2021, nov_2021, dec_2021, jan_2022, feb_2022, mar_2022, apr_2022, may_2022)
```

-- Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020  
*In the newer datasets 2021 >, birthyear and gender are already removed -Chad*  

```{r}
all_trips <- all_trips |>
  select(-c(start_lat, start_lng, end_lat, end_lng))
```


## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS  

-- Inspect the new table that has been created  

```{r data cleaning}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

*Note: Some of these steps below will not be needed, as I selected more recent data than what is used in the example, and these files seem to be already modified as mentioned in the steps below. -Chad *  

### There are a few problems we will need to fix:  
1. In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
2. The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year that provide additional opportunities to aggregate the data.
3. We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
4. There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.  

-- In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"  
-- Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature  
*N.B.: "Level" is a special property of a column that is retained even if a subset does not contain any values from a specific level *  
-- Begin by seeing how many observations fall under each usertype  

```{r}
table(all_trips$member_casual)
```

-- Add columns that list the date, month, day, and year of each ride
This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level  
*[date formats](https://www.statmethods.net/input/dates.html) more on date formats in R found at that link *  

```{r add date columns}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

-- Add a "ride_length" calculation to all_trips (in seconds)  
--- [difftime](https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html)  

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

-- Inspect the structure of the columns  

```{r}
str(all_trips)
```

-- Convert "ride_length" from Factor to numeric so we can run calculations on the data  

```{r ride length factor conversion}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

-- Remove "bad" data  
The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative. We will create a new version of the dataframe (v2) since data is being removed
[delete or drop rows](https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/)  

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
```


## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS  

-- Descriptive analysis on ride_length (all figures in seconds)  

```{r descriptive analysis}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
summary(all_trips_v2$ride_length)
```

--- Compare members and casual users  

```{r aggregate data}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

--- See the average ride time by each day for members vs casual users  

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

--- Notice that the days of the week are out of order. Let's fix that.  

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

--- Now, let's run the average ride time by each day for members vs casual users  

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

--- analyze ridership data by type and weekday  

```{r analyze ridership}
all_trips_v2 |>
  mutate(weekday = wday(started_at, label = TRUE)) |>  #creates weekday field using wday()
  group_by(member_casual, weekday) |>  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) |> 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

-- Let's visualize the number of rides by rider type  
*I added a title and the new pipe syntax -Chad *  

```{r first plot}
all_trips_v2 |> 
  mutate(weekday = wday(started_at, label = TRUE)) |> 
  group_by(member_casual, weekday) |> 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) |> 
  arrange(member_casual, weekday)  |> 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title="Number of Riders by Rider Type", subtitle=paste0("Data from June 2021 through May 2022"))
```

-- Let's create a visualization for average duration  
*I added a title and the new pipe syntax -Chad *  

```{r}
all_trips_v2 |> 
  mutate(weekday = wday(started_at, label = TRUE)) |> 
  group_by(member_casual, weekday) |> 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) |> 
  arrange(member_casual, weekday)  |> 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title="Average Ride Duration by Rider Type", subtitle=paste0("Data from June 2021 through May 2022"))
```


## STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS  

-- Create a csv file that we will visualize in Excel, Tableau, or my presentation software  
--- N.B.: This file location is for a Mac. If you are working on a PC, change the file location accordingly (most likely "C:\Users\YOUR_USERNAME\Desktop\...") to export the data. [You can read more here:](https://datatofish.com/export-dataframe-to-csv-in-r/)  

```{r export summary data}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = '/Users/chadsisk/Education/Coursera/Google/DataAnalytics/Case Studies/Cyclistic/avg_ride_length.csv')
```

